Mapping Cancer: the Potential of Cartograms and Alternative Map Displays

### 2020-05-25 literature.Rmd
Final changes for resubmission

# 1. Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments, not-for-profits and the media are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations for the geographical units, especially for privacy protection, but also for political and policy purposes. Examples of typical geographical units include states, provinces, local government areas, and post/zip codes. Counts and incidence rates are often obtained by state health registries. For this data privacy and ethics are key concerns, even when combined to small area statistics.
This type of data is routinely collected for public health reasons and may be made available to the general public as a service to the community.

@@

This paper considers the current visualization techniques to communicate statistics to the public and their applications to cancer statistics. Alternative approaches are posed because they may be more effective than contemporary techniques. The limitations of the visualization methods, highlighting the differences and historic use of these displays is discussed.

@@

|Measure | Details | Fig. |
|:--------------------|:---------------------------------|:----|
|1. Cancer incidence counts| the number of new cases occurring in a specified period and geographic area [@Globocan]| |
|2. Incidence Rate (per 100,000) | The frequency of cases relative to a fixed population size, usually 100,000 people, within a specific period of time [@Globocan] | e, g |
|3. Incidence Ratio | The incidence rate in a region divided by the average incidence rate for all of the regions| c |
|4. Age-Standardized Incidence Rate (per 100,000) | The weighted mean of the age-specific incidence rates; the weights are taken from population distribution of the standard population. The most frequently used standard population is the World Standard Population. [@Globocan] | b, d |
|5. Age-Standardized Relative Risk | The relative risks represent the risk of an area (ward) relative to average risk, adjusted for age, deprivation and chance fluctuations due to small numbers. [@EnvEnglandWales2] | a, f |

@@

| Fig. | Atlas | Data | Statistic | Source |
|----|-----------|-------------|-----------|--------------|
| a | The Environment and Health<br> Atlas of England and Wales<br> [@EnvEnglandWales2] | Sex: Women<br> Year: 2010<br> Cancer: Lung | Relative risk | Office for National Statistics<br> (ONS) (England) and from the<br> Welsh Cancer Intelligence and<br> Surveillance Unit (WCISU) [1a](http://www.envhealthatlas.co.uk/homepage/) |
| b | Globocan 2018: Estimated Cancer<br> Incidence, Mortality and <br> <br> prevalence Worldwide<br> [@Globocan] | Sex: Both<br> Year: 2018<br> Cancer: All invasive | Age standardized<br> incidence rates<br> (per 100,000) | World Health Organization's<br> International Agency for<br> Research on Cancer [1b](https://www.iarc.fr/<br>)  |
| c | Atlas of Cancer in Queensland<br>  [@QLDcancerAtlas] | Sex: Males<br> Year: 1998 - 2007<br> Cancer: Lung | Relative incidence<br> ratio | Queensland Cancer Registry [1c](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/) |
| d | Australian Cancer Atlas<br> [@ACA] | Sex: Males<br> Year: 2010 - 2014<br> Cancer: Lung | Average standardised<br> incidence rate | Australian Cancer Atlas [1d](https://atlas.cancer.org.au/) |
| e | United States Cancer<br> Statistics: An Interactive<br> Cancer Statistics Website<br>  | Sex: Both<br> Year: 2016<br> Cancer: All types | Incidence rate per<br> 100,000 | With data from state cancer<br> registries: <br> Centers for Disease Control and Prevention [1e](https://www.cdc.gov/) |
| f | Map of Cancer Mortality<br> Rates in Spain<br> [@cancerSpain] | Sex: Both<br> Year: 2004 - 2008<br> Cancer: All types | Relative risk | <br> Map of cancer mortality rates<br> in Spain [1f](https://elpais.com/elpais/2014/10/06/media/1412612722_141933.html) |
| g | Atlas of Childhood<br> Cancer in Ontario<br> [@OntarioPediatric] | Sex: Both<br> Year: 1995 - 2004<br> Cancer: All types | Incidence rate per<br> 100,00 | <br> The Pediatric Oncology Group<br> of Ontario Networked Information<br> System [1g](https://www.pogo.ca/research-data/pogonis-childhood-cancer-database/data-requests/) |

@@

Choropleth maps colour each geographic unit to allow map users to measure the value of the statistic [@EI]. Map users contrast the colours in neighbouring areas to understand the spatial distribution. The ColorBrewer system [@CB] and viridis [@viridis] palettes provide effective colour schemes for qualitative, sequential and diverging data. When communicating information using colour, a map creator should use a scheme with perceptually uniform color spaces that match equal steps in data space with equal steps in the colour space [@PUCS], a linear color gradient is appropriate for incidence counts and rates, but may not be appropriate for ratio measures. It is possible to allow for data to progress uniformly in both positive and negative directions from a mid point, such as the mean of the data. These diverging colour schemes pair two sequential schemes that use a common light colour at the mid point, each sequential scheme progresses to a dark hue at the extreme value in each direction [@DCS].
The use of borders and backgrounds, and their colours, can also change the appearance of the colors representing the value of the statistics [@CB]. These supports can be used to implement a reference point in the colour scheme as well as orient users to the geographic regions.

@@

<!-- statistics -->
Many statistics are commonly used in cancer displays. It is common to see incidence rates, or relative rates which measure how far a region is above or below the average. The purpose of using an age standardised relative rate is, perhaps the desire to pinpoint the areas that need attention because they have higher than expected rates. 
<!--However, we lose the incidence rate information and thus interpretability. --> A region might be much higher than average, but it may not be close to a health concern, because all regions have a low incidence. Supplementary materials can allow map users to recognise when this occurs.

>>>>

# 1. Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments, not-for-profits and the media are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Counts and incidence rates are often obtained by state health registries, as data privacy and ethics are key concerns, even when combined to small area statistics. Presenting these statistics requires aggregating individual observations for the geographical units, especially for privacy protection, but also for political and policy purposes. Examples of typical geographical units include states, provinces, local government areas, and post/zip codes. 
This type of data is routinely collected for public health reasons and may be made available to the general public as a service to the community.

@@

This paper considers the current visualization techniques to communicate statistics to the public and their applications to cancer statistics. Alternative approaches are posed because they may be more effective than contemporary techniques. The differences and historic use of these displays is discussed, highlighting the potential and the limitations of the visualization methods.

@@

|Measure | Details | Fig. |
|:--------------------|:---------------------------------|:----|
|1. Cancer incidence counts| the number of new cases occurring in a specified period and geographic area [@Globocan]| |
|2. Incidence Rate (per 100,000) | The frequency of cases relative to a fixed population size, usually 100,000 people, within a specific period of time [@Globocan] | e, g |
|3. Incidence Ratio | The incidence rate in a region divided by the average incidence rate for all of the regions| c |
|4. Age-Standardized Incidence Rate (per 100,000) | The weighted mean of the age-specific incidence rates; the weights are taken from population distribution of the standard population. The most frequently used standard population is the World Standard Population. [@Globocan] | b |
|4. Age-Standardized Incidence Ratio | The ratio of the observed Age-Standardized Incidence Rate, to the expected number of cases. Having adjusted for differences in population and differences in age structure of the population within each area [@DACA] | d |
|5. Age-Standardized Relative Risk | The relative risks represent the risk of an area (ward) relative to average risk, adjusted for age, deprivation and chance fluctuations due to small numbers. [@EnvEnglandWales2] | a, f |

@@

| Fig. | Atlas | Data | Statistic | Source |
|----|-----------|-------------|-----------|--------------|
| a | The Environment and Health<br> Atlas of England and Wales<br> [@EnvEnglandWales2] | Sex: Women<br> Year: 2010<br> Cancer: Lung | Relative risk | Office for National Statistics<br> (ONS) (England) and from the<br> Welsh Cancer Intelligence and<br> Surveillance Unit (WCISU) [1a](http://www.envhealthatlas.co.uk/homepage/) |
| b | Globocan 2018: Estimated Cancer<br> Incidence, Mortality and <br> <br> prevalence Worldwide<br> [@Globocan] | Sex: Both<br> Year: 2018<br> Cancer: All invasive | Age standardized<br> incidence rates<br> (per 100,000) | World Health Organization's<br> International Agency for<br> Research on Cancer [1b](https://www.iarc.fr/<br>)  |
| c | Atlas of Cancer in Queensland<br>  [@QLDcancerAtlas] | Sex: Males<br> Year: 1998 - 2007<br> Cancer: Lung | Relative incidence<br> ratio | Queensland Cancer Registry [1c](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/) |
| d | Australian Cancer Atlas<br> [@ACA] | Sex: Males<br> Year: 2010 - 2014<br> Cancer: Lung | Age-standardised<br> incidence ratio | Australian Cancer Atlas [1d](https://atlas.cancer.org.au/) |
| e | United States Cancer<br> Statistics: An Interactive<br> Cancer Statistics Website<br>  | Sex: Both<br> Year: 2016<br> Cancer: All types | Incidence rate per<br> 100,000 | With data from state cancer<br> registries: <br> Centers for Disease Control and Prevention [1e](https://www.cdc.gov/) |
| f | Map of Cancer Mortality<br> Rates in Spain<br> [@cancerSpain] | Sex: Both<br> Year: 2004 - 2008<br> Cancer: All types | Relative risk | <br> Map of cancer mortality rates<br> in Spain [1f](https://elpais.com/elpais/2014/10/06/media/1412612722_141933.html) |
| g | Atlas of Childhood<br> Cancer in Ontario<br> [@OntarioPediatric] | Sex: Both<br> Year: 1995 - 2004<br> Cancer: All types | Incidence rate per<br> 100,00 | <br> The Pediatric Oncology Group<br> of Ontario Networked Information<br> System [1g](https://www.pogo.ca/research-data/pogonis-childhood-cancer-database/data-requests/) |

@@

Choropleth maps colour each geographic unit to allow map users to measure the value of the statistic [@EI]. Map users contrast the colours in neighbouring areas to understand the spatial distribution. Pickle's [@USMA] suggestions for choropleth map displays include directions to categorise the statistic presented on the map according to percentiles. A colour scheme that communicates high and low rates can be useful in displays of cancer statistics, using a double ended colour scheme such as those provided by the ColorBrewer system [@CB] and viridis [@viridis] palettes provide effective colour schemes for qualitative, sequential and diverging data. When communicating information using colour, a map creator should use a scheme with perceptually uniform color spaces that match equal steps in data space with equal steps in the colour space [@PUCS]. It is possible to allow for data to progress uniformly in both positive and negative directions from a mid point, such as the mean of the data. These diverging colour schemes pair two sequential schemes that use a common light colour at the mid point, each sequential scheme progresses to a dark hue at the extreme value in each direction [@DCS], it prefereable that the scheme use darker or warmer colors for higher cancer rates [@USMA]. A linear color gradient is appropriate for incidence counts and rates. The linear gradient can also be transformed using the log scale, to show appropriate colours for ratio measures. The Australian Cancer Atlas example in Figure 1d implements dark red for areas with an SIR value with a risk level more than 50% (eg. SIR = 1.5) above the Australian average. Areas were coloured dark blue if they had an SIR value below the inverse of the risk value specified (eg. ≈ 0.67). The use of borders and backgrounds, and their colours, can also change the appearance of the colors representing the value of the statistics [@CB]. These supports can be used to implement a reference point in the colour scheme as well as orient users to the geographic regions.

@@

<!-- statistics -->
Many statistics are commonly used in cancer displays. It is common to see incidence rates, or ratios which displays how far a region is above or below the average. The purpose of using an age standardised ratio is, perhaps the desire to pinpoint the areas that need attention because they have higher than expected rates. Ratios may negatively impact the interpretability of the actual rates of incidence or mortality. However this impact is offset by the value of seeing the relativity of the values in reference to the mean, especially when the mean value is also given on the display. This helps to put individual areas in perspective, 
<!--However, we lose the incidence rate information and thus interpretability. --> as a region might have a value higher than the average, but it may not be a health concern if all regions have a low incidence rate. Supplementary materials such as displaying the mean, or a distribution plot, can allow map users to recognise when this occurs.
